Supplementary Material

### 2025-01-15 paper.qmd
state of the WD when submitted to SR

**Authors**: : Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.

@@

The aim of this work was to develop and validate a novel non-linear model to characterize RR interval (RRi) dynamics throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches. Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi dynamics, with parameter estimation achieved via Hamiltonian Monte Carlo. Sobol sensitivity analysis identified baseline RRi ($\alpha$) and recovery proportion ($c$) as the primary drivers of variability, underscoring their critical roles in autonomic regulation and resilience. Validation against real-world RRi data demonstrated robust model performance, accurately reflecting autonomic recovery and exercise-induced fluctuations. By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring, as well as athletic contexts for optimizing performance and recovery. Taken together, these findings highlight the model's ability to provide precise, physiologically relevant assessments of autonomic function, paving the way for its use in personalized health monitoring and performance optimization across diverse populations.

@@

In this context, the study of R-R intervals (RRi), defined as the time intervals between heartbeats, and its link with exercise has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices]. Hence, being able to analyze the temporal dynamics of RRi in response to exercise can provide valuable insights into the mechanisms by which the cardiovascular system adapts to physical stressors, such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].

@@

Modeling the RRi behavior has been traditionally approached by leveraging linear regression and time-series analysis [@lian2011simple]. However, this oppose significant challenges, like oversimplifying overall exercise dynamics without capturing the intricacies of exercise-induced intricate transitions in RRi, especially under intense exertion and recovery phases [@gronwald2019effects]. More recently, advanced non-linear approaches have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated;@lee2023real;@berrahou2024arrhythmia]. However, many of the more advance alternatives fails to generate physiologically meaningful model parameters, without a direct link to biological processes [@gronwald2019non].

@@

Considering the current state of the art, there is a clear need to develop a non-linear model that accurately represents RRi's non-linear transitions during exercise and recovery, where the estimation of clinically significant parameters is compelling. Such a model would need to offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy]. This model would need to be complex enough to capture the non-linear, exercise-induced, cardiovascular dynamics. But, at the same time, simple enough to be able to provide practical and significant model parameters related to observed physiological processes.

@@

The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity]. Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.

@@

The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package (v2.21.0), which employs the Stan probabilistic programming language. Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, which were kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes, which were targeted at a minimum of 1,000 for each parameter [@brms]. Trace plots were inspected to confirm stable mixing. These diagnostics collectively confirmed reliable posterior estimates for each parameter.

@@

To enhance the exploration of parameter space, we performed a two-stage analysis: An individual-level estimation of parameter values that were then used for the estimation population-level parameters.

@@

$$
y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-\beta_i \cdot c}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
$${#eq-indiv-level}
where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.

@@

where $\mu_{k}$ is the group-level mean for parameter $k$, $b_{k,i}$ represents the subject-level random effect for subject $i$ on parameter $k$ and $\sigma_k^2$ is the residual variance for parameter $k$. The subject-level effects $b_{k,i}$ were assumed to be distributed as $b_{k,i} \sim \mathcal{N}(0,\sigma^2)$, with $\sigma$ being the standard error of the subject-level effect.

This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates into subject and population-level model parameters.

@@

The primary performance metrics included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject. Bootstrap resampling across each metric was performed to estimate the mean performance of the model and corresponding quantile-based 95% CI.

@@

To assess the sensitivity of model parameters influencing RRi over time, we implemented a Sobol sensitivity analysis using Monte Carlo simulations. Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise [@zhang2015sobol].

To compute $S_{\text{ind}}$, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs in total). For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes. The 95% CI parameter values estimated from HMC-NUTS were then used as input ranges for $S_{\text{ind}}$ computation. Finally, the mean values of $S_{\text{ind}}$ over the 20-min time span for each model parameter were estimated and reported, next with their corresponding 95% CI using normal approximation based on estimated standard errors (SE).

@@

RRi dynamics in response to exercise tends to follow a U-shaped form. The initial decrease in RRi is associated with exercise onset, related to an increase in heart rate. After exercise cessation, an opposite increase in RRi is observed, associated with the cardiovascular recovery phase. In both cases, the drop and recovery phases occurs at different rates, some individuals experience a quick recovery in RRi after exercise, however, in some others this slope is less steep. Additionally, the new baseline reached following exercise cessation is often below the RRi baseline before exercise.

These hallmarks of RRi dynamics in response to exercise, highlights the complex and non-linear behavior of the cardiovascular response in the context of both rest and exercise conditions. An example RRi record data is shown in [figure 1](#fig-rri-example).

@@

The process of deriving the final equation for modeling RRi fluctuations was guided by an iterative exploration of mathematical functions capable of capturing the observed dynamics. Initially, exponential and logarithmic functions were considered due to their simplicity and wide applicability in describing temporal changes. Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset, while logarithmic functions were explored for their capacity to describe asymptotic behaviors observed in some physiological variables.

However, neither approach successfully reproduced the non-linear and bidirectional nature of the RRi fluctuations. Exponential functions, while effective at modeling monotonic decay or growth, could not account for the observed sigmoidal transitions. Similarly, logarithmic functions, with their inherent monotonicity, failed to represent the plateauing behavior seen in real-world data.

To address these limitations, we shifted to logistic functions, which inherently model sigmoidal transitions. Logistic functions introduce parameters for growth rate and inflection point, allowing for precise control over the shape and timing of the transition between dynamic states. By using two coupled logistic functions, one to represent the initial decrease in RRi and a second, inverted logistic function to describe the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.

This approach provided a biologically plausible representation, with parameters that directly correspond to identifiable physiological features, such as the rate of adaptation and recovery, the time to peak response, and the extent of deviation from baseline. The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured both the qualitative and quantitative aspects of RRi dynamics.

@@

In this equation, $a_1$ represents the asymptotic value approached by the logistic function, which can be either positive (indicating an increase) or negative (indicating a decrease). For $f_1(t)$, this parameter is specified as $\beta$, indicating the absolute change in RRi at the onset of exercise. In contrast, for $f_2(t)$, $a_1$ is parametrized as $-c \cdot \beta$, where $c$ denotes the proportion of change relative to the initial drop indicated by $\beta$. This parametrization ensures that, after the initial decline, the second logistic function facilitates the return of RRi toward the baseline value $\alpha$.

@@

The sample used to assess RRi dynamics consists of a group of 272 subjects selected from a local community of elderly individuals. The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)

@@

Initial graphical exploration of RRi dynamics (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated the exercise-induced cardiovascular stress. However, greater variability across individuals in post-exercise recovery can be observed.

@@

Given the observed parameters that reproduced the observed RRi patterns in exercise and rest conditions, priors were chosen based on physiological constraints and graphical visualization of standardized RRi data. Hence, ensuring identifiability of model parameters by constraining the parameter space to plausible values, in order to improve model convergence and parameter exploration. The prior distributions were defined as follows:

@@

Once subject-level RRi data was fitted using the proposed in @eq-indiv-level, a population-parameter value was estimated based on the proposed group-level methodology. The estimated parameter values can be seen in [Table 2](#tbl-parameters)

@@

Relative performance metrics, estimated through bootstrapped resampling, suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data. This is equivalent to a 32.6 ms in the RRi scale (CI~95%~[30.01, 35.77]). Additionally, the bootstrapped R^2^ indicates that the model explains 0.868 (CI~95%~[0.834, 0.895]) of the total variance observed in RRi.

Residuals analysis showed that the estimated partial correlation function (ACF) from the model residuals indicates a correlation among non-explained errors greater than 0.1 up to the 5th lag. However, the partial ACF is significant (CI-wise) and strictly positive or negative up to the second lag. Correlations among model residuals against other time indices remained insignificant (see [Figure 6](#fig-error-rates)).

@@

Individual perturbation of each parameter highlighted that RRi dynamics are sensitive to the baseline RRi parameter, $\alpha$. Conversely, the rate parameters for the initial decay during exercise, $\lambda$, and the recovery post-exercise, $\phi$, show lower sensitivity, suggesting that they are not primary sources of variation in predicted RRi trajectories when assessed in isolation. The results of the sensitivity analysis can be seen in [Table 3](#tbl-sens-params).

@@

To the best of our knowledge, this study represents the first attempt to develop a non-linear model specifically designed to continuously capture RR interval dynamics across a complete rest-exercise-rest protocol. Previous studies have either focused on aggregate HRV indices or utilized simplified linear or exponential models, which are insufficient to describe the complex, non-stationary transitions observed during and after exercise [@molkkari2020online]. By employing a combination of logistic functions, our model uniquely accounts for the gradual shifts in autonomic regulation denoted by RRi dynamics, offering a detailed and physiologically relevant representation of cardiac dynamics. This continuous modeling framework allows for the integration of both exercise-induced RRi decline and post-exercise recovery within a single unified structure, bridging a critical gap in the current literature. Such an approach not only advances our understanding of cardiovascular responses but also opens new avenues for real-time monitoring and intervention in both clinical and athletic settings.

The proposed model demonstrates a precise capacity to reproduce RRi dynamics, with its combination of logistic functions, capturing the key transitions of cardiac response, the initial decline during exercise and the subsequent recovery. This design accommodates the inherent non-linearity and non-stationarity of RRi dynamics, overcoming the limitations of linear models and exponential functions commonly used in prior studies [@silva2019exponential; @mongin2022decrease].

Compared to previous research, our findings align with efforts made on capturing nonlinear dynamics in HRV to understand cardiac responses during exercise [@gronwald2019effects]. Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness, supporting the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv]. However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi dynamics that enhances interpretability and application across diverse fitness levels and exercise intensities.

@@

Unlike prior research that aggregates HRV measures or applies simple decay models, our approach directly models RRi changes, offering richer physiological insight. For instance, commonly utilized exponential decay models for post-exercise recovery are used, however they fail to incorporate the transition dynamics observed during exercise itself [@sahoo2024unanticipated]. By integrating both exercise and recovery phases, our model provides a more comprehensive view of autonomic regulation.

Moreover, the sensitivity of parameters such as $\lambda$ (decay rate) and $(\phi$ (recovery rate) was found to be relatively low, suggesting that the model is robust to variability in these rates while remaining sensitive to key physiological parameters ($\alpha$ and $c$). This robustness makes it suitable for both individualized monitoring and population-level analyses, offering versatility in its application across different use cases.

@@

This model demonstrates significant potential for practical applications in both clinical and athletic settings. In clinical contexts, it could aid in tailoring cardiovascular rehabilitation protocols by monitoring autonomic recovery in real-time, ensuring safe and effective exercise regimens for at-risk populations [@carrasco2024hrv]. This aligns with previous research, which highlight the importance of individualizing rehabilitation programs to optimize recovery [@takahashi2020signs; @rizvi2023enhancing; @carrasco2024hrv].

In athletic settings, the model could guide training strategies, particularly for interval training, where determining optimal recovery periods is crucial. Similar findings suggest that precise monitoring of RRi dynamics can prevent overtraining and enhance performance [@hebisz2022heart; @nuuttila2024monitoring]. The model's ability to integrate real-time data from wearable devices further enhances its applicability in dynamic, uncontrolled environments, enabling field-based monitoring and feedback [@zimatore2020recurrence].

@@

Additionally, environmental and psychological factors, such as temperature, stress, or sleep quality, were not explicitly considered in this study. Future work could integrate these variables into the model, enhancing its robustness and applicability across varied real-world scenarios. This aligns with calls for more integrative modeling approaches in cardiovascular research [@takahashi2020signs; @hebisz2022heart; @nuuttila2024monitoring].

@@

In summary, this study presents a novel non-linear model for RRi dynamics, capturing the complex and transient autonomic responses during rest-exercise-recovery protocols, overcoming the limitations of traditional autonomic metrics. By identifying baseline RRi and recovery proportion as the dominant contributors to variability, the model emphasizes their critical roles in reflecting autonomic regulation and resilience. Validated across a cohort of elderly participants, the model demonstrates robust performance in real-time cardiovascular assessments, offering significant potential for clinical applications such as rehabilitation and monitoring in at-risk populations, as well as athletic contexts like fatigue management and performance optimization. While the model's applicability is currently constrained by its focus on elderly individuals, future validation in younger cohorts and under diverse environmental conditions will enhance its generalizability and utility. This work establishes a foundational framework for advancing personalized cardiovascular health monitoring and intervention.

@@

Conceptualization, MC-A; Data curation, MC-A; Investigation, MC-A; Methodology, MC-A, NMD; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E, DM-C; Writing–review & editing, MC-A, CN-E, DM. All authors have read and agreed to the published version of the manuscript.

@@

![](figures/example-data.png)
[**Figure 1**]{#fig-rri-example}. Example data of RRi recordings of 6 different subjects over a 20 minute rest-exercise-recovery protocol in a sample of elderly individuals. The subject-level data shows the inter-individual variability of RRi dynamics in response exercised-induced cardiovascular stress, with similar behavior and trajectories of recovery over time.

@@

![](figures/2d-density-kernel.png)
[**Figure 3**]{#fig-2d-kernel-density}. (A) Mean and SD from each of the subject's RRi recordings, used for the standardization process. (B) 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol. Darker colors indicate greater probability density. The contrary can be said about lighter colors.

@@

![](figures/prior-model.png)
[**Figure 4**]{#fig-prior-sim}. (A) Simulated standardized RRi dynamics based on prior parameter distributions, illustrating predicted RRi responses to exercise. Shaded areas represent 95%, 80%, and 60% quantile CI, offering insight into expected physiological variability across parameters. (B) Prior distributions and 95% CI, used to generate prior predictions, based on physiological constraints and graphical visualization of standardized RRi data.

@@

![](figures/group-level-estimates.png)
[**Figure 5**]{#fig-group-level-estimates}. (A) Posterior probability distributions of the expectation for each population-parameter estimates ($E[\theta]$) with quantile-based 95% CI. (B) Transformed rate parameters into a percentage scale using the $1 - \exp(\theta)$ transformation.

@@

![](figures/error-estimates.png)
[**Figure 6**]{#fig-error-rates}. Individual-level performance metrics. (A) Bootstrapped MAPE and RMSE, as metrics of relative and absolute model deviance from observed RRi. (B) Individual-level estimates of model performance and the relationship between them. (C) Partial autocorrelation function (ACF) of model residuals with corresponding quantile-based CI.

@@

[**Table 1**]{#tbl-sample-characteristics}. Sample characteristics from which, continuous RRi monitoring data was collected during a rest-exercise-rest protocol. Data is presented as Mean ± standard deviation (SD). SBP, systolic blood pressure; DBP, diastolic blood pressure; MAP, mean arterial pressure; PP, pulse pressure; BMI, body mass index.

>>>>

**Authors**: Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.

@@

This work aimed to develop and validate a novel non-linear model to characterize RR interval (RRi) dynamics throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches. Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi dynamics, with parameter estimation achieved via Hamiltonian Monte Carlo. Sobol sensitivity analysis identified baseline RRi ($\alpha$) and recovery proportion ($c$) as the primary drivers of variability, underscoring their critical roles in autonomic regulation and resilience. Validation against real-world RRi data demonstrated robust model performance, accurately reflecting autonomic recovery and exercise-induced fluctuations. By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring in athletic contexts to optimize performance and recovery. These findings highlight the model's ability to provide precise, physiologically relevant assessments of autonomic function, paving the way for its use in personalized health monitoring and performance optimization across diverse populations.

@@

In this context, the study of R-R intervals (RRi), defined as the time intervals between heartbeats and their link with exercise, has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices]. Hence, analyzing the temporal dynamics of RRi in response to exercise can provide valuable insights into how the cardiovascular system adapts to physical stressors, such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].

@@

Modeling the RRi behavior has traditionally been approached by leveraging linear regression and time-series analysis [@lian2011simple]. However, this oppose significant challenges, like oversimplifying overall exercise dynamics without capturing the intricacies of exercise-induced intricate transitions in RRi, especially under intense exertion and recovery phases [@gronwald2019effects]. More recently, advanced non-linear approaches have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated;@lee2023real;@berrahou2024arrhythmia]. However, many more advanced alternatives fail to generate physiologically meaningful model parameters without directly linking to biological processes [@gronwald2019non].

@@

Considering the current state of the art, there is a clear need to develop a non-linear model that accurately represents RRi's non-linear transitions during exercise and recovery, where the estimation of clinically significant parameters is compelling. Such a model would need to offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy]. This model must be complex enough to capture the non-linear, exercise-induced cardiovascular dynamics. But, simultaneously, it has to be simple enough to provide practical and significant model parameters related to observed physiological processes.

@@

The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity]. Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, which is capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.

@@

The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package (v2.21.0), which employs the Stan probabilistic programming language. Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes, targeted at a minimum of 1,000 for each parameter [@brms]. Trace plots were inspected to confirm stable mixing. These diagnostics collectively confirmed reliable posterior estimates for each parameter.

@@

To enhance the exploration of parameter space, we performed a two-stage analysis: We assessed parameter values at the individual level, which we then used to estimate  population-level parameters.

@@

$$
y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-c_i \cdot \beta_i}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
$${#eq-indiv-level}
where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.

@@

where $\mu_{k}$ is the group-level mean for parameter $k$, $b_{k,i}$ represents the subject-level random effect for the subject $i$ on parameter $k$ and $\sigma_k^2$ is the residual variance for the parameter $k$. The subject-level effects $b_{k,i}$ were assumed to be distributed as $b_{k,i} \sim \mathcal{N}(0,\sigma^2)$, with $\sigma$ being the standard error of the subject-level effect.

This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates of subject- and population-level model parameters.

@@

The primary performance metrics, estimated for each subject, included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject. Bootstrap resampling across each metric was performed to estimate the mean performance of the model and corresponding quantile-based 95% CI.

@@

We implemented a Sobol sensitivity analysis using Monte Carlo simulations to assess the sensitivity of model parameters influencing RRi over time. Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise [@zhang2015sobol].

To compute $S_{\text{ind}}$, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs). For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes. The 95% CI parameter values estimated from HMC-NUTS were then used as input ranges for $S_{\text{ind}}$ computation. Finally, the mean values of $S_{\text{ind}}$ over the 20-minute time span for each model parameter were estimated and reported, with their corresponding 95% CI using normal approximation based on estimated standard errors (SE).

@@

RRi dynamics in response to exercise tends to follow a U-shaped form. The initial decrease in RRi is associated with exercise onset, related to an increased heart rate. After exercise cessation, an opposite increase in RRi is observed, which is associated with the cardiovascular recovery phase. In both cases, the drop and recovery phases occurs at different rates; some individuals experience a quick recovery in RRi after exercise; however, in some others, this slope is less steep. Additionally, the new baseline reached following exercise cessation is often below the RRi baseline before exercise.

These hallmarks of RRi dynamics in response to exercise, highlight the complex and non-linear behavior of the cardiovascular response in the context of both rest and exercise conditions. [Figure 1](#fig-rri-example) shows an example of RRi record data.

@@

The process of deriving the final equation for modeling RRi fluctuations was guided by an iterative exploration of mathematical functions capable of capturing the observed dynamics. Initially, exponential and logarithmic functions were considered due to their simplicity and wide applicability in describing temporal changes. Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset. In contrast, logarithmic functions were explored for their capacity to describe asymptotic behaviors observed in some physiological variables.

However, neither approach successfully reproduced the non-linear and bidirectional nature of the RRi fluctuations. While effective at modeling monotonic decay or growth, exponential functions could not account for the observed sigmoidal transitions. Similarly, logarithmic functions, with their inherent monotonicity, failed to represent the plateauing behavior seen in real-world data.

We shifted to logistic functions to address these limitations, which inherently model sigmoidal transitions. Logistic functions introduce parameters for growth rate and inflection point, allowing for precise control over the shape and timing of the transition between dynamic states. By using two coupled logistic functions, one to represent the initial decrease in RRi and a second inverted logistic function describing the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.

This approach provided a biologically plausible representation, with parameters that directly correspond to identifiable physiological features, such as the rate of adaptation and recovery, the time to peak response, and the extent of deviation from baseline. The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured the qualitative and quantitative aspects of RRi dynamics.

@@

In this equation, $a_1$ represents the asymptotic value the logistic function approaches, which can be either positive (indicating an increase) or negative (indicating a decrease). For $f_1(t)$, this parameter is specified as $\beta$, indicating the absolute change in RRi at the onset of exercise. In contrast, for $f_2(t)$, $a_1$ is parametrized as $-c \cdot \beta$, where $c$ denotes the proportion of change relative to the initial drop indicated by $\beta$. This parametrization ensures that, after the initial decline, the second logistic function facilitates the return of RRi toward the baseline value $\alpha$.

@@

The sample used to assess RRi dynamics consists of 272 subjects selected from a local community of elderly individuals. The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)

@@

An initial graphical exploration of RRi dynamics (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated with exercise-induced cardiovascular stress. However, greater variability across individuals in post-exercise recovery can be observed.

@@

Given the parameters that reproduced the observed RRi patterns in exercise and rest conditions, priors were chosen based on physiological constraints and the graphical visualization of standardized RRi data. Hence, ensuring the identifiability of model parameters by constraining the parameter space to plausible values to improve model convergence and parameter exploration. The prior distributions were defined as follows:

@@

Once subject-level RRi data was fitted using the proposed model in @eq-indiv-level, a population-parameter value was estimated based on the proposed group-level methodology. The estimated parameter values can be seen in [Table 2](#tbl-parameters)

@@

Estimated through bootstrapped resampling, relative performance metrics suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data. This is equivalent to a 32.6 ms in the RRi scale (CI~95%~[30.01, 35.77]). Additionally, the bootstrapped R^2^ indicates that the model explains 0.868 (CI~95%~[0.834, 0.895]) of the total variance observed in RRi.

Residuals analysis showed that the estimated partial correlation function (ACF) from the model residuals indicates a correlation among non-explained errors greater than 0.1 up to the 5th lag. However, the partial ACF is significant (CI-wise) and strictly positive or negative until the second lag. Correlations among model residuals against other time indices remained insignificant (see [Figure 6](#fig-error-rates)).

@@

Individual perturbation of each parameter highlighted that RRi dynamics are sensitive to the baseline RRi parameter, $\alpha$. Conversely, the rate parameters for the initial decay during exercise, $\lambda$, and the recovery post-exercise, $\phi$, show lower sensitivity, suggesting that they are not primary sources of variation in predicted RRi trajectories when assessed in isolation. The results of the sensitivity analysis are in [Table 3](#tbl-sens-params).

@@

To the best of our knowledge, this study represents the first attempt to develop a non-linear model specifically designed to capture RRi dynamics across a complete rest-exercise-rest protocol continuously. Previous studies have either focused on aggregate HRV indices or utilized simplified linear or exponential models, which are insufficient to describe the complex, non-stationary transitions observed during and after exercise [@molkkari2020online]. By employing a combination of logistic functions, our model uniquely accounts for the gradual shifts in autonomic regulation denoted by RRi dynamics, offering a detailed and physiologically relevant representation of cardiac dynamics. This continuous modeling framework integrates exercise-induced RRi decline and post-exercise recovery within a single unified structure, bridging a critical gap in the current literature. Such an approach advances our understanding of cardiovascular responses and opens new avenues for real-time monitoring and intervention in both clinical and athletic settings.

The proposed model demonstrates a precise capacity to reproduce RRi dynamics. With its combination of logistic functions, it captures the key transitions of cardiac response, the initial decline during exercise, and the subsequent recovery. This design accommodates the inherent non-linearity and non-stationarity of RRi dynamics, overcoming the limitations of linear models and exponential functions commonly used in prior studies [@silva2019exponential; @mongin2022decrease].

Compared to previous research, our findings align with efforts to capture nonlinear dynamics in HRV to understand cardiac responses during exercise [@gronwald2019effects]. Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness. This supports the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv]. However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi dynamics that enhances interpretability and application across diverse fitness levels and exercise intensities.

@@

Unlike prior research that aggregates HRV measures or applies simple decay models, our approach directly models RRi changes, offering richer physiological insight. For instance, commonly utilized exponential decay models for post-exercise recovery are used but fail to incorporate the transition dynamics observed during exercise itself [@sahoo2024unanticipated]. By integrating exercise and recovery phases, our model provides a more comprehensive view of autonomic regulation.

Moreover, the sensitivity of parameters such as $\lambda$ (decay rate) and $(\phi$ (recovery rate) was found to be relatively low, suggesting that the model is robust to variability in these rates while remaining sensitive to key physiological parameters ($\alpha$ and $c$). This robustness makes it suitable for individualized monitoring and population-level analyses, offering versatility in its application across different use cases.

@@

This model demonstrates significant potential for practical applications in clinical and athletic settings. In clinical contexts, it could aid in tailoring cardiovascular rehabilitation protocols by monitoring autonomic recovery in real-time, ensuring safe and effective exercise regimens for at-risk populations [@carrasco2024hrv]. This aligns with previous research, which highlights the importance of individualizing rehabilitation programs to optimize recovery [@takahashi2020signs; @rizvi2023enhancing; @carrasco2024hrv].

The model could guide training strategies in athletic settings, particularly for interval training, where determining optimal recovery periods is crucial. Similar findings suggest that precise monitoring of RRi dynamics can prevent overtraining and enhance performance [@hebisz2022heart; @nuuttila2024monitoring]. The model's ability to integrate real-time data from wearable devices further enhances its applicability in dynamic, uncontrolled environments, enabling field-based monitoring and feedback [@zimatore2020recurrence].

@@

Additionally, this study did not explicitly consider environmental and psychological factors, such as temperature, stress, or sleep quality. Future work could integrate these variables into the model, enhancing its robustness and applicability across varied real-world scenarios. This aligns with calls for more integrative modeling approaches in cardiovascular research [@takahashi2020signs; @hebisz2022heart; @nuuttila2024monitoring].

@@

In summary, this study presents a novel non-linear model for RRi dynamics, capturing the complex and transient autonomic responses during rest-exercise-recovery protocols, overcoming the limitations of traditional autonomic metrics. The model emphasizes their critical roles in reflecting autonomic regulation and resilience by identifying baseline RRi and recovery proportion as the dominant contributors to variability. Validated across a cohort of elderly participants, the model demonstrates robust performance in real-time cardiovascular assessments, offering significant potential for clinical applications such as rehabilitation and monitoring in at-risk populations and athletic contexts like fatigue management and performance optimization. While the model's applicability is currently constrained by its focus on elderly individuals, future validation in younger cohorts and under diverse environmental conditions will enhance its generalizability and utility. This work establishes a foundational framework for advancing personalized cardiovascular health monitoring and intervention.

@@

Conceptualization, MC-A; Data curation, MC-A; Investigation, MC-A, DM-C; Methodology, MC-A, NMD; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E, DM-C; Writing–review & editing, MC-A, CN-E, DM. All authors have read and agreed to the published version of the manuscript.

@@

![](figures/example-data.png)
[**Figure 1**]{#fig-rri-example}. Example data of RRi recordings of 6 subjects over a 20-minute rest-exercise-recovery protocol in a sample of elderly individuals. The subject-level data shows the inter-individual variability of RRi dynamics in response to exercised-induced cardiovascular stress, with similar behavior and recovery trajectories over time.

@@

![](figures/2d-density-kernel.png)
[**Figure 3**]{#fig-2d-kernel-density}. (A) Mean and SD from each subject's RRi recordings were used for the standardization process. (B) 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol. Darker colors indicate greater probability density. The contrary can be said about lighter colors.

@@

![](figures/prior-model.png)
[**Figure 4**]{#fig-prior-sim}. (A) Simulated standardized RRi dynamics based on prior parameter distributions, illustrating predicted RRi responses to exercise. Shaded areas represent 95%, 80%, and 60% quantile CI, offering insight into expected physiological variability across parameters. (B) Prior distributions and 95% CI were used to generate prior predictions based on physiological constraints and graphical visualization of standardized RRi data.

@@

![](figures/group-level-estimates.png)
[**Figure 5**]{#fig-group-level-estimates}. (A) Posterior probability distributions of the expectation for each population-parameter estimate ($E[\theta]$) with quantile-based 95% CI. (B) Transformed rate parameters into a percentage scale using the $1 - \exp(\theta)$ transformation.

@@

![](figures/error-estimates.png)
[**Figure 6**]{#fig-error-rates}. Individual-level performance metrics. (A) Bootstrapped MAPE and RMSE are metrics of relative and absolute model deviance from observed RRi. (B) Individual-level estimates of model performance and the relationship between them. (C) Partial autocorrelation function (ACF) of model residuals with corresponding quantile-based CI.

@@

[**Table 1**]{#tbl-sample-characteristics}. Sample characteristics from which continuous RRi monitoring data was collected during a rest-exercise-rest protocol. Data is presented as Mean ± standard deviation (SD). SBP, systolic blood pressure; DBP, diastolic blood pressure; MAP, mean arterial pressure; PP, pulse pressure; BMI, body mass index.
